---
title: "Feedback for Week 3 Homework"
author: "Juanjo Medina"
date: "22 February 2018"
output: slidy_presentation
---

## Start session reading data and packages

```{r cars, echo = TRUE}
library(ggplot2)

ncovr <- read.csv("NAT.csv")
```

## Faceting

```{r pressure}
ggplot(data = ncovr, aes(x = HR90)) + 
  geom_histogram() + 
  facet_wrap(~SOUTH)
```

##Boxplots

```{r}
ggplot(data = ncovr,
       aes(x = as.factor(SOUTH), y = HR90)) + 
  geom_boxplot()
```

##Histograms

```{r}
ggplot(ncovr, aes(x = HR90, fill = as.factor(SOUTH))) +
  geom_histogram(position = "identity", alpha = 0.4)
```

##Using the logarithmic scale

```{r}
ggplot(ncovr, aes(x = log10(HR90+1), fill = as.factor(SOUTH))) +
  geom_histogram(position = "identity", alpha = 0.4)
```

##Density plots

```{r}
ggplot(ncovr, aes(x = log10(HR90 + 1), fill = as.factor(SOUTH))) + 
  geom_density(alpha = .3)
```

##Violint plots

```{r}
ggplot(data = ncovr,
       aes(x = as.factor(SOUTH), y = log10(HR90+1))) + 
  geom_violin() 
```

For details see: Hintze, J. L., Nelson, R. D. (1998) Violin Plots: A Box Plot-Density Trace Synergism. The American Statistician 52, 181-184

##Show me the points

```{r}
ggplot(data = ncovr,
       aes(x = as.factor(SOUTH), y = log10(HR90+1))) + 
  geom_violin() +
  geom_jitter(width = 0.2, alpha = 0.2)
```

##Adjusting properties of the violins

```{r}
ggplot(data = ncovr,
       aes(x = as.factor(SOUTH), y = log10(HR90+1))) + 
  geom_violin(scale="count") +
  geom_jitter(width = 0.2, alpha = 0.2)

```

```{r}
table(ncovr$SOUTH)
```

##Scatterplots

```{r}
ggplot(ncovr, aes(x = UE70, y = HR70)) +
  geom_point(alpha=.4) +
  geom_smooth(colour="red", size=1, se=FALSE)
```

##Conditioning on a third variable

```{r}
ggplot(ncovr, aes(x = UE70, y = HR70, colour= as.factor(SOUTH))) +
  geom_point(alpha=.1) +
  geom_smooth(size=1.2, se=FALSE)
```

##Only scratching the surface

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(sf)
library(tmap)
shp_name <- "ncovr/NAT.shp"
ncovr_map <- st_read(shp_name)
```


```{r}
tm_shape(ncovr_map) + 
  tm_fill("HR70", title = "Homicide Rate 1990", style="quantile", palette= "Reds") +
  tm_borders(alpha = 0.1) +
  tm_layout(legend.position = c("left", "bottom"), legend.title.size = 0.8)
```


##Further resources

-Hadley Wickham ggplot2 page: http://ggplot2.tidyverse.org/index.html

-Kieran Helay book: http://socviz.co

